Bayesian Sparsification of Deep $\cplx$-valued networks

### 2020-02-05 paper/paper.tex
scattered edits

\textit{Stochastic Gradient Variational Bayes} (SGVB), proposed by \citet{kingma_auto-encoding_2014},
is a differentiable unbiased Monte-Carlo estimator of \eqref{eq:elbo_general}, which
makes it possible to employ stochastic gradient-based optimization methods to solve
\eqref{eq:variational-progam}. If $
  \omega \sim q_{\theta}
$ is equivalent in distribution to $
  \omega = g(\varepsilon; \theta)
$ for some non-parametric random variable $
    \varepsilon \sim p_\varepsilon
$ and differentiable $
  g(\varepsilon; \theta)
$, then the SGVB is
\begin{equation}  \label{eq:sgvb}
  \widetilde{\mathcal{L}}(\theta; \phi, \lambda)
    = - KL(q_{\theta} \| \pi_{\lambda})
      % + \frac{N}{\lvert B \rvert} \sum_{i\in B}
        % \log p(z_i \mid \omega_i)b
      + \frac1{L} \sum_{l=1}^L
        \log p_{\phi}(D \mid g(\varepsilon_{l}; \theta))
        % \log p_{\phi}(D \mid \omega_{l})
        %   \Big\vert_{\omega_{l} = g(\varepsilon_{l}; \theta)}
      % =
      % - KL(q_{\theta} \| \pi)
      % + N \mathbb{E}_{B \sim \{D\}_M \otimes q_\theta^M}
        % \hat{\mathbb{E}}_{z,\omega \sim B}
          % \log p(z \mid \omega)
    \,, 
\end{equation}
where $
  (\varepsilon_{l})_{l=1}^L
$ is an iid sample from $p_\varepsilon$. Intractable KL-divergence terms can be replaced
by differentiable unbiased sampling estimators \citep{kingma_auto-encoding_2014}.
%
For the purposes of this study, we assume that the observed data is independent, and
use minibatch version of \eqref{eq:sgvb} with $L=1$:
\begin{equation}  \label{eq:elbo}
  \widetilde{\mathcal{L}}(\theta; \phi, \lambda)
    = - KL(q_{\theta} \| \pi_{\lambda})
      + \frac{N}{M} \sum_{i=1}^M
        \log p_{\phi}(z_i \mid g(\varepsilon_i; \theta))
    \,,
\end{equation}
for a random subsample $(z_i)_{i=1}^M$ from $D$ and iid $
  (\varepsilon_i)_{i=1}^M \sim p_\varepsilon
$.
%
% (DSVI) titsias-lazaro.tex \todo{DSVI} \citep{titsias_doubly_2014}

@@

Complex networks are, in general, non-holomorphic, i.e. not $\cplx$-differentiable, which is
exacerbated by the loss being real-valued. Indeed a $\cplx$-differentiable real-valued function is
necessarily trivial by Cauchy-Riemann conditions. This issue is dealt with by employing Wirtinger,
or $\cplx\real$ calculus, which generalizes holomorphic calculus to non-holomorphic functions
of $\cplx$-argument \citep{adali_complex-valued_2011,boeddeker_computation_2019}. It satisfies
product and chain rules, respects complex conjugation and linearity, but most importantly the
$\cplx\real$-differential of a function on $\cplx$ coincides with its classical differential as
a function on $\real^2$. Essentially, it allows straightforward retrofiting of $\cplx$-valued
networks into existing $\real$ deep learning auto-differentiation frameworks. It was considered
as a basis for $\cplx$ version of backpropagation by \citet{benvenuto_complex_1992} and
\citet{trabelsi_deep_2017}, (see sec.~\ref{sub:wirtinger_calculus} for details).
% \todo{read on holomorphic nets}

@@

The local reparameterization trick provides both faster computations, and better statistical
efficiency of the SGVB gradient estimator \citep{wang_fast_2013,kingma_variational_2015}.
Suppose that weights $W \in \cplx^{n\times m}$ of a dense linear layer are drawn from a
mean field $\cplx$-Gaussian approximate posterior
\begin{equation}  \label{eq:c-gauss-vi-general}
  W_{ij}
    \sim \mathcal{N}^{\cplx} \bigl(
      \mu_{ij}, \Sigma_{ij}, C_{ij}
    \bigr)
  \,,
\end{equation}
where mean, \textit{variance} and \textit{relation} are given by matrices $
  \mu, \Sigma, C \in \cplx^{n\times m}
$ with $\Sigma_{ij} \in \real$, and $
  \lvert C_{ij} \rvert^2 \leq \Sigma_{ij}
$. The property \eqref{eq:cn-affine} and independence of elements of $W$ imply that for any
$x \in \cplx^m$ and $b \in \cplx^n$ the components of $y = W x + b$ are independent with
\begin{equation}  \label{eq:cplx-gauss-trick}
  y_i
    \sim \mathcal{N}^{\cplx}
      \Bigl(
        % e_i^\top \mu x + b_i,
        b_i + \sum_{j=1}^m \mu_{ij} x_j,
        \, \sum_{j=1}^m \Sigma_{ij} \lvert x_j \rvert^2,
        \, \sum_{j=1}^m C_{ij} x_j^2
      \Bigr)
    \,.
\end{equation}
The trick requires three matrix operations: $\mu x + b$, $\Sigma \lvert x \rvert^2$ and
$C x^2$ with the modulus and power applied elementwise.

@@

We consider two priors for Bayesian dropout: \verify{log-uniform} and $\cplx$-Gaussian empirical
Bayes \citep{kingma_variational_2015,molchanov_variational_2017,kharitonov_variational_2018}.
For a fully factorized approximation $q(\omega)$ and factorized prior belief $\pi(\omega)$,
the divergence term in ELBO \eqref{eq:elbo} is
\begin{equation}  \label{eq:elbo-general-kl-div}
  KL(q \| \pi)
    % = \mathbb{E}_{\oemga \sim q(\oemga)} \log \tfrac{q(\oemga)}{\pi(\oemga)}
    % = \sum_{ij} \mathbb{E}_{\omega_{ij} \sim q(\omega_{ij})}
    %   \log \tfrac{q(\omega_{ij})}{\pi(\omega_{ij})}
    % = \sum_{ij} KL\bigl( q(\omega_{ij}) \| \pi(\omega_{ij}) \bigr)
    = - \sum_{ij}
        \mathbb{H}(q(\omega_{ij}))
        + \mathbb{E}_{q(\omega_{ij})} \log{\pi(\omega_{ij})}
    \,.
\end{equation}
We use approximation \eqref{eq:c-gauss-vi} and omit subscripts ${ij}$ for brevity.

@@

Consider the following parameterization of $W$: $
  W = \mu \odot \xi
$ where $
  \xi_{ij} \sim \mathcal{N}(1, \alpha_{ij})
$, yet $\mu \in \cplx^{n \times m}$. This case corresponds to inference regarding
relevance coefficients $\xi$ rather than the parameters themselves after. Under this
parameterization
\begin{equation}  \label{eq:real-scal-gauss-vi}
  W_{ij}
    \sim \mathcal{N}^{\cplx} \bigl(
      \mu_{ij},
      \, \alpha_{ij} \lvert \mu_{ij} \rvert^2,
      \, \alpha_{ij} \mu_{ij}^2
    \bigr)
  \,,
\end{equation}
i.e. a degenerate $\cplx$-Gaussian distribution, which implies by \eqref{eq:cplx-gauss-trick}
\begin{equation}  \label{eq:real-scal-gauss-trick}
  y_i
    \sim \mathcal{N}^{\cplx}\bigl(
      % e_i^\top \mu x + b,
      b_i + \sum_{j=1}^m \mu_{ij} x_j,
      \, \sum_j \alpha_{ij} \lvert x_{ij} \mu_{ij}\rvert^2,
      \, \sum_j \alpha_{ij} (x_{ij} \mu_{ij})^2
    \bigr)
    \,.
\end{equation}
The KL-divergence term for this approximation coincides with \eqref{eq:improper-kl-div-real},
which is approximated by \citet{molchanov_variational_2017}. The major drawback of this method
is inapplicability of the additive noise reparameterization in \eqref{eq:real-scal-gauss-trick},
which disentangles the gradient with respect to $\mu$ from the local Gaussian noise.

>>>>

\textit{Stochastic Gradient Variational Bayes} (SGVB), proposed by \citet{kingma_auto-encoding_2014},
is a differentiable unbiased Monte-Carlo estimator of \eqref{eq:elbo_general}, which
makes it possible to employ stochastic gradient-based optimization methods to solve
\eqref{eq:variational-progam}. If $
  \omega \sim q_{\theta}
$ is equivalent in distribution to $
  \omega = g(\varepsilon; \theta)
$ for some non-parametric random variable $
    \varepsilon \sim p_\varepsilon
$ and differentiable $
  g(\varepsilon; \theta)
$, then the SGVB is
\begin{equation}  \label{eq:sgvb}
  \widetilde{\mathcal{L}}(\theta; \phi, \lambda)
    = - KL(q_{\theta} \| \pi_{\lambda})
      % + \frac{N}{\lvert B \rvert} \sum_{i\in B}
        % \log p(z_i \mid \omega_i)b
      + \frac1{L} \sum_{l=1}^L
        \log p_{\phi}(D \mid g(\varepsilon_{l}; \theta))
        % \log p_{\phi}(D \mid \omega_{l})
        %   \Big\vert_{\omega_{l} = g(\varepsilon_{l}; \theta)}
      % =
      % - KL(q_{\theta} \| \pi)
      % + N \mathbb{E}_{B \sim \{D\}_M \otimes q_\theta^M}
        % \hat{\mathbb{E}}_{z,\omega \sim B}
          % \log p(z \mid \omega)
    \,, 
\end{equation}
where $
  (\varepsilon_{l})_{l=1}^L
$ is an iid sample from $p_\varepsilon$. Intractable KL-divergence terms can be replaced
by differentiable unbiased sampling estimators \citep{kingma_auto-encoding_2014}.
%
For the purposes of this study, we assume that the observed data is independent, and
use minibatch version of \eqref{eq:sgvb} with $L=1$:
\begin{equation}  \label{eq:elbo}
  \widetilde{\mathcal{L}}(\theta; \phi, \lambda)
    = - KL(q_{\theta} \| \pi_{\lambda})
      + \frac{N}{M} \sum_{i=1}^M
        \log p_{\phi}(z_i \mid g(\varepsilon_i; \theta))
    \,,
\end{equation}
for a random subsample $(z_i)_{i=1}^M$ from $D$ and iid $\varepsilon_i \sim p_\varepsilon$.
%
% (DSVI) titsias-lazaro.tex \todo{DSVI} \citep{titsias_doubly_2014}

@@

Complex networks are, in general, non-holomorphic, i.e. not $\cplx$-differentiable, which is
exacerbated by the loss being real-valued. Indeed a $\cplx$-differentiable real-valued function is
necessarily trivial by Cauchy-Riemann conditions. This issue is dealt with by employing Wirtinger,
or $\cplx\real$ calculus, which generalizes holomorphic calculus to non-holomorphic functions
of $\cplx$-argument \citep{adali_complex-valued_2011,boeddeker_computation_2019}. It satisfies
product and chain rules, respects complex conjugation and linearity, but most importantly the
$\cplx\real$-differential of a function on $\cplx$ coincides with its classical differential as
a function on $\real^2$. Essentially, it allows straightforward retrofiting of $\cplx$-valued
networks into existing $\real$ deep learning auto-differentiation frameworks. It was considered
as a basis for $\cplx$ version of backpropagation by \citet{benvenuto_complex_1992} and
\citet{trabelsi_deep_2017}, (see appendix~\ref{sub:wirtinger_calculus} for details).
% \todo{read on holomorphic nets}

@@

The local reparameterization trick provides both faster computations, and better statistical
efficiency of the SGVB gradient estimator \citep{wang_fast_2013,kingma_variational_2015}.
Suppose that weights $W \in \cplx^{n\times m}$ of a linear layer are drawn from a mean
field $\cplx$-Gaussian approximation
\begin{equation}  \label{eq:c-gauss-vi-general}
  W_{ij}
    \sim \mathcal{N}^{\cplx} \bigl(
      \mu_{ij}, \Sigma_{ij}, C_{ij}
    \bigr)
  \,,
\end{equation}
where mean, \textit{variance} and \textit{relation} are given by matrices $
  \mu, \Sigma, C \in \cplx^{n\times m}
$ with $\Sigma_{ij} \in \real$, and $
  \lvert C_{ij} \rvert^2 \leq \Sigma_{ij}
$. The property \eqref{eq:cn-affine} and independence of elements of $W$ imply that for any
$x \in \cplx^m$ and $b \in \cplx^n$ the components of $y = W x + b$ are independent with
distributions
\begin{equation}  \label{eq:cplx-gauss-trick}
  y_i
    \sim \mathcal{N}^{\cplx}
      \Bigl(
        % e_i^\top \mu x + b_i,
        b_i + \sum_{j=1}^m \mu_{ij} x_j,
        \, \sum_{j=1}^m \Sigma_{ij} \lvert x_j \rvert^2,
        \, \sum_{j=1}^m C_{ij} x_j^2
      \Bigr)
    \,.
\end{equation}
The trick requires three matrix operations: $\mu x + b$, $\Sigma \lvert x \rvert^2$ and
$C x^2$ with the modulus and power applied elementwise.

@@

We consider two priors for Bayesian dropout: \verify{log-uniform} and $\cplx$-Gaussian empirical
Bayes \citep{kingma_variational_2015,molchanov_variational_2017,kharitonov_variational_2018}.
For a fully factorized approximation $q(\omega)$ and factorized prior belief $\pi(\omega)$,
the divergence term in ELBO \eqref{eq:elbo} is
\begin{equation}  \label{eq:elbo-general-kl-div}
  KL(q \| \pi)
    % = \mathbb{E}_{\oemga \sim q(\oemga)} \log \tfrac{q(\oemga)}{\pi(\oemga)}
    % = \sum_{ij} \mathbb{E}_{\omega_{ij} \sim q(\omega_{ij})}
    %   \log \tfrac{q(\omega_{ij})}{\pi(\omega_{ij})}
    % = \sum_{ij} KL\bigl( q(\omega_{ij}) \| \pi(\omega_{ij}) \bigr)
    = - \sum_{ij}
        \mathbb{H}(q(\omega_{ij}))
        + \mathbb{E}_{q(\omega_{ij})} \log{\pi(\omega_{ij})}
    \,.
\end{equation}
We omit subscripts ${ij}$ for brevity in the next sectons.

@@

Consider the following parameterization of $W$: $
  W = \mu \odot \xi
$ where $
  \xi_{ij} \sim \mathcal{N}(1, \alpha_{ij})
$, yet $\mu \in \cplx^{n \times m}$. This case corresponds to inference regarding
relevance coefficients $\xi$ rather than the parameters themselves after. Each weight
under this parameterization is an independent degenerate $\cplx$-Gaussian distribution
\begin{equation}  \label{eq:real-scal-gauss-vi}
  W_{ij}
    \sim \mathcal{N}^{\cplx} \bigl(
      \mu_{ij},
      \, \alpha_{ij} \lvert \mu_{ij} \rvert^2,
      \, \alpha_{ij} \mu_{ij}^2
    \bigr)
  \,,
\end{equation}
which implies by \eqref{eq:cplx-gauss-trick} the following:
\begin{equation}  \label{eq:real-scal-gauss-trick}
  y_i
    \sim \mathcal{N}^{\cplx}\bigl(
      % e_i^\top \mu x + b,
      b_i + \sum_{j=1}^m \mu_{ij} x_j,
      \, \sum_j \alpha_{ij} \lvert x_{ij} \mu_{ij}\rvert^2,
      \, \sum_j \alpha_{ij} (x_{ij} \mu_{ij})^2
    \bigr)
    \,.
\end{equation}
The KL-divergence term for this approximation coincides with \eqref{eq:improper-kl-div-real},
which is approximated by \citet{molchanov_variational_2017}. The major drawback of this method
is inapplicability of the additive noise reparameterization in \eqref{eq:real-scal-gauss-trick},
which disentangles the gradient with respect to $\mu$ from the local noise.
